Automated Metadata Generation for Fish Specimen Image Collections {\footnotesize \thanks{NSF Grant ???} }

### 2021-03-31 paper/paper_new.tex

\title{Automated Metadata Generation for Fish Specimen Image Collections%*\\
    {\footnotesize \thanks{NSF Grant ???} }}
%do we add grant now, or wait til acceptance. i think we need to ultimately remove any identifying info., i think bgnn is likely okay; but, perhaps mention metadata research center (MRC) need to be removed?. if so, i kind of like to put it back in for final version, if accepted. it helps with getting funding ... you are both listed on the MRC page. we should check on the identifying info. removed, or not.
%for the record: Research supported by NSF OAC #1940233 and #1940322.

@@

This limitation is magnified when trying get a handle on the actual quality of the digitized specimen. Descriptive oriented standards support search and retrieval, and the biodiversity community advocated for data fitness standards \cite{chapman2020developing}. Even so, metadata properties used to record image quality is severely limited and generally missing. This point is addressed in detail in \cite{leipzig2021biodiversity}, justifying the capture of 22 metadata properties supporting selection of high quality images for the development of a bgnn. Although, the manual generation of this metadata is costly in terms of labor and time.
%will add cites above, depending on paper length.

@@

\subsection{Automatic Metadata Generation}
Technical metadata, such as camera settings and temporal information (date and time) are automatically generated during a digitization sequence, following standards such as XX. Some of this metadata may be useful in for selecting a machine learning sample. A researcher may desire images captured chiefly with a certain aperture setting. Even so, the majority of automatically captured technical metadata associated with digitized specimens is insufficient for computational research, leaving researchers to reply on manually generated descriptive metadata, which itself is sparse.

Automatic metadata generation of descriptive bibliographic data has been a research focus for close to 20 years \cite{liddy2002automatic,greenberg2004metadata,cardinaels2005automating}. Researchers have applied support vector machine (SVM) approaches \cite{han2003automatic}, and associated networks to address sparse and incomplete metadata \cite{rodriguez2009automatic}, and various successes are integrated into day-to-day workflows.

%\cite{heidorn2008automatic} speciment labels \cite{manso2004automatic} carteogrpahic info.

@@

\subsubsection{Threshold Adjustment}
The first calculation in the pixel analysis process determines the cutoff intensity between what constitutes the foreground (i.e.~the fish) and background of the image.
Initially, the calculation is based on the bounding box and mask generated by
\verb|detectron|. Specimen images are read in as gray scale, and pixels in the image are treated as unsigned integers between 0 and 255.
Otsu thresholding, a technique that maximizes the variance between the 
foreground and background intensities, is used to compute an initial cutoff value between foreground and background. 
While the Otsu value occasionally generates an accurate mask as is, usually
the contrast between foreground and background is low and much of the lighter parts of the fish (such as its tail fin) are marked as background.

@@

\subsubsection{Fallback} The pixel analysis process occasionally fails. This can occur if certain flood fill operations behave unexpectedly, or if the image is too washed out or otherwise atypical for the thresholding process to work correctly. In the event this happens, the original mask and bounding box generated by \verb|detectron| is used for metadata generation.
\subsection{Metadata Generation}
Including what has already been discussed, the following metadata properties are being currently being generated:

\begin{table*}[t]
    \centering
    \caption{Metadata properties}
    \label{tab:properties}
    \begin{tabular}{cccp{0.5\linewidth}}
        \toprule
        \textbf{Property} & \textbf{Association} & \textbf{Type} & \textbf{Explanation}\\
        \midrule
        \verb|has_fish| & Overall Image & Boolean & Whether a fish was found in the image.\\
        \verb|fish_count| & Overall Image & Integer & The quantity of fish present.\\
        \verb|has_ruler| & Overall Image & Boolean & Whether a ruler was found in the image.\\
        \verb|ruler_bbox| & Overall Image & 4 Tuple & The bounding box of the ruler (if found).\\
        \verb|scale| & Overall Image & Float & The scale of the image in \(\frac{\mathrm{pixels}}{\mathrm{cm}}\).\\
        \verb|bbox| & Per Fish & 4 Tuple & The top left and bottom right coordinates of what define the bounding box for a fish.\\
        \verb|background.mean| & Per Fish & Float & The mean intensity of the background within a given fish's bounding box.\\
        \verb|background.std| & Per Fish & Float & The standard deviation of the background within a given fish's bounding box.\\
        \verb|foreground.mean| & Per Fish & Float & The mean intensity of the foreground within a given fish's bounding box.\\
        \verb|foreground.std| & Per Fish & Float & The standard deviation of the foreground within a given fish's bounding box.\\
        \verb|centroid| & Per Fish & 4 Tuple & The centroid of a given fish's bitmask.\\
        \verb|primary_axis| & Per Fish & 2D Vector & The unit length primary axis (eigenvector) for the bitmask of a given fish.\\
        \verb|clock_value| & Per Fish & Integer & A fish's primary axis converted into an integer ``clock value'' between 1 and 12.\\
        \verb|length| & Per Fish & Float & The length of a fish in \(\frac{\mathrm{pixels}}{\mathrm{cm}}\).\\
        \verb|mask| & Per Fish & 2D Matrix & The bitmask of a fish in 0's and 1's.\\
        \verb|pixel_analysis_failed| & Per Fish & Boolean & Whether the pixel analysis process failed for a given fish. If \verb|true|, \verb|detectron|'s mask and bounding box were used for metadata generation.\\
        \verb|score| & Per Fish & Float & The percent confidence score output by \verb|detectron| for a given fish.\\
        \verb|has_eye| & Per Fish & Boolean & Whether an eye was found for a given fish.\\
        \verb|eye_center| & Per Fish & 2 Tuple & The centroid of a fish's eye.\\
        \verb|side| & Per Fish & String & The side (i.e.\ \verb|'left'| or \verb|'right'|) of the fish that is facing the camera (dependent on finding its eye).\\
      \bottomrule
\end{tabular}
\end{table*}

\verb|has_fish|, \verb|fish_count|, \verb|has_ruler|, \verb|ruler_bbox|, \verb|background.*|, \verb|foreground.*|, \verb|bbox|, \verb|mask|, \verb|score|, and \verb|has_eye| have already been discussed. The remaining properties are determined as follows:
\subsubsection{centroid and eye\_center}
Centroids are provided for masks/bounding boxes generated by \verb|detectron|, and since we do not recalculate the mask of fish eyes we can use that value directly for \verb|eye_center|.

Since we recalculate the mask of the fish, we must recalculate its centroid as well. This can be done via
\begin{equation}
    (\bar{x}, \bar{y}) = (\mathrm{round}(\frac{M_{10}}{M_{00}}), \mathrm{round}(\frac{M_{01}}{M_{00}}))
\end{equation}
where \(M_{00}\) is the pixel area of the fish's blob, \(M_{10}\) is the sum of all the \(x\) values of blob pixels, and \(M_{01}\) is the sum of all the \(y\) values of blob pixels.
\subsubsection{side}
Determining which \verb|side| of the fish is visible is predicated on finding its eye. If an eye is found, the sign of the \(x\) component of the vector from the centroid of the fish to the centroid of the eye specifies which side is up: negative for left and positive for right. This assumes the fish was photographed vertically, which is essentially always the case for all image collections our group has worked on.
\subsubsection{primary\_axis and clock\_value}
The \verb|primary_axis| of a fish can be calculated by taking the covariance\todo{I could explain this in more detail if necessary} of its blob in \(x\) and \(y\), which yields its principle eigenvector. This can be directly assigned to the property. If an eye is present, we specify this axis in the direction of the eye in relation to the fish's centroid.

Our team encoded this information as a ``clock value'' between 1 and 12 when recording it by hand. To convert \verb|principal_axis| to \verb|clock_value|, the sign of \(x\) and \(y\) on the principal axis are used to determine which Cartesian quadrant the fish is angles into relative to its centroid. Depending on this quadrant, we then dot product the principal axis with either \([-1,0]\), \([0,-1]\), \([1,0]\) or \([0,1]\) which correspond to 9, 6, 3 and ``0'' o'clock respectively. The resulting radian value is then converted to a polar displacement in clock value space, and added to the comparative clock value used in the dot product. This then gives the fish's clock value from 0 to \(11.\overline{9}\). Before recording as \verb|clock_value| in the output, the value is rounded to the nearest integer, and in the event of a 0 final result it is replaced with 12.
\subsubsection{scale and length}
\(\frac{\mathrm{pixels}}{\mathrm{inch}}\) can be calculated by measuring the distance between the digits 2 and 3 found on the ruler by \verb|detectron|. Converting this to \(\frac{\mathrm{pixels}}{\mathrm{cm}}\) gives the \verb|scale| metadata property as reported in the output.

For the fish \verb|length| property, it is necessary to determine the furthest points from the centroid of the fish in each direction. Since fish are normally measured in a straight line from their snout down the middle of their trunk, we project every pixel onto the major axis of the fish (as a line through its centroid) before checking the distance. This projection is done by intersecting the line through the point along the tangent of the principal axis and the aforementioned centroid--principal axis line. After checking every pixel, measuring the distance between the two furthest projected points gives the length of the fish in pixels. Multiplying this by \verb|scale| gives the fish \verb|length| in centimeters.
\begin{comment}
\subsection{Consideration}
Just gonna throw some caveats in here, might be its own section might not
\begin{itemize}
    \item Program falls back to using detectron mask and bbox for metadata generation in the event the mask generation/pixel analysis fails. Failure is defined as never finding a cohesive flood filled mask that covers more than 10\% of the detectron bbox initially. Can provide a statistic on how often this occurs.
\end{itemize}
\end{comment}
\section{Results}
Technicians employed by our team have manually generated the 22 metadata properties deemed crucial to the overall BGNN project~\cite{Leipzig2021.01.28.428644} for a large number of INHS images. \(20,699\) total entries were created by 13 technicians that spanned \(8,398\) unique images. Of these \(8,398\) images, \(7,244\) were both not part of the \verb|detectron| training set and meet our current admissibility criteria for what \verb|detectron| and the pixel analysis process can reliably handle. We ran the metadata extraction program on these \(7,244\) images. These properties did not include information on image scale, fish length, or fish bounding boxes. For these properties, a random sample of 50 specimens from the set of \(7,244\) were analyzed by hand for comparison.

@@

\subsection{Fish Detection}
All images in the INHS dataset contain exactly one fish. For \(7,209\) of the specimen images, one fish was detected. For 25 of the images, 2 fish were detected, 3 fish were detected for 3 images, and for 7 images no fish were found. In the case of greater than 1 fish, 9 of the 28 contained tags which overlapped the fish and were themselves labeled as a second fish. Of the remaining 17, \verb|detectron| erroneously labeled the fish as two separate fish objects, or labeled a subsection of the fish a second time.

\subsection{Ruler Detection}
For all but 2 of the \(7,244\) images \verb|detectron| was able to find the ruler.\todo{Not sure if I can phrase it that way or not since I haven't looked at all the images to see if the ruler itself was found or something else was labeled a ruler} For an additional 56 images, the ruler itself was found but the ``2'' and/or the ``3'' on it was not, so a scale calculation could not be performed.

@@

@@

\subsection{Mask and Bounding Box}
\begin{figure}[H]
  \centering
  \includegraphics[width=0.49\linewidth]{images/011196_pred}
  \includegraphics[width=0.49\linewidth]{images/011196_mask}
  \includegraphics[width=0.49\linewidth]{images/52399_pred}
  \includegraphics[width=0.49\linewidth]{images/52399_mask}
  \includegraphics[width=0.49\linewidth]{images/56885_pred}
  \includegraphics[width=0.49\linewidth]{images/56885_mask}
  \caption{Examples of masks and bounding boxes from detectron (left) and pixel analysis (right).}
\end{figure}
Fish bounding boxes were calculated for all \(7,237\) images in which a fish was found. All but 263 of these were generated via pixel analysis, with those 263 falling back to the original \verb|detectron| bounding box. Of the 50 images reviewed manually for these properties, all mask and bounding boxes were correctly placed on/around the location of the fish. However, a number of them lacked portions of lightly colored tails and/or fins. Specifically, 22 mask and bounding boxes covered the entire fish, 16 missed some of the tail, and 12 missed most or all of the tail.

>>>>

\title{Automated Metadata Generation for Fish Specimen Image Collections%*\\
    {\footnotesize \thanks{NSF Grant ???} }}
%do we add grant now, or wait til acceptance. i think we need to ultimately remove any identifying info., e.g., bgnn, and even mention of metadata research center. 
%for the record: Research supported by NSF OAC #1940233 and #1940322.

@@

This limitation is magnified, when trying get a handle on the actual quality of the digitized specimen. Descriptive oriented standards support search and retrieval, but metadata properties used to record image quality is severely limited and generally missing. This point is addressed in detail in \cite{leipzig2021biodiversity}, justifying the capture of 22 metadata properties supporting selection of high quality images for the development of a bio-generated neural network.
%will add cites above, depending on paper length.

@@

\subsection{Automatic Metadata Generation}
Metadata researchers are aware of challenges posted by manual metadata generation, both in terms of time,  //Technical metadata capturing image lighting, camera, date of digitization and other information are automatically generated, but 
%jane working here, i figured out how to scope this, I think! I need to get outside now, but back later.

@@

\begin{table*}[t]
    \centering
    \caption{Metadata properties}
    \label{tab:properties}
    \begin{tabular}{cccp{0.5\linewidth}}
        \toprule
        \textbf{Property} & \textbf{Association} & \textbf{Type} & \textbf{Explanation}\\
        \midrule
        \verb|has_fish| & Overall Image & Boolean & Whether a fish was found in the image.\\
        \verb|fish_count| & Overall Image & Integer & The quantity of fish present.\\
        \verb|has_ruler| & Overall Image & Boolean & Whether a ruler was found in the image.\\
        \verb|ruler_bbox| & Overall Image & 4 Tuple & The bounding box of the ruler (if found).\\
        \verb|scale| & Overall Image & Float & The scale of the image in \(\frac{\mathrm{pixels}}{\mathrm{cm}}\).\\
        \verb|bbox| & Per Fish & 4 Tuple & The top left and bottom right coordinates of the bounding box for a fish.\\
        \verb|background.mean| & Per Fish & Float & The mean intensity of the background within a given fish's bounding box.\\
        \verb|background.std| & Per Fish & Float & The standard deviation of the background within a given fish's bounding box.\\
        \verb|foreground.mean| & Per Fish & Float & The mean intensity of the foreground within a given fish's bounding box.\\
        \verb|foreground.std| & Per Fish & Float & The standard deviation of the foreground within a given fish's bounding box.\\
        \verb|centroid| & Per Fish & 4 Tuple & The centroid of a given fish's bitmask.\\
        \verb|primary_axis| & Per Fish & 2D Vector & The unit length primary axis (eigenvector) for the bitmask of a given fish.\\
        \verb|clock_value| & Per Fish & Integer & Fish's primary axis converted into an integer ``clock value'' between 1 and 12.\\
        \verb|length| & Per Fish & Float & The length of a fish in \(\frac{\mathrm{pixels}}{\mathrm{cm}}\).\\
        \verb|mask| & Per Fish & 2D Matrix & The bitmask of a fish in 0's and 1's.\\
        \verb|pixel_analysis_failed| & Per Fish & Boolean & Whether the pixel analysis process failed for a given fish. If \verb|true|, \verb|detectron|'s mask and bounding box were used for metadata generation.\\
        \verb|score| & Per Fish & Float & The percent confidence score output by \verb|detectron| for a given fish.\\
        \verb|has_eye| & Per Fish & Boolean & Whether an eye was found for a given fish.\\
        \verb|eye_center| & Per Fish & 2 Tuple & The centroid of a fish's eye.\\
        \verb|side| & Per Fish & String & The side (i.e.\ \verb|'left'| or \verb|'right'|) of the fish that is facing the camera (dependent on finding its eye).\\
      \bottomrule
\end{tabular}
\end{table*}

\subsubsection{Threshold Adjustment}
The first calculation in the pixel analysis process determines the cutoff intensity between what constitutes the foreground (i.e.~the fish) and background of the image.
Initially, the calculation is based on the bounding box and mask generated by
\verb|detectron|. Specimen images are read in as gray scale, and pixels in the image are treated as unsigned integers between 0 and 255.
Otsu thresholding \cite{Otsu1979ATS}, a technique that maximizes the variance between the 
foreground and background intensities, is used to compute an initial cutoff value between foreground and background. 
While the Otsu value occasionally generates an accurate mask as is, usually
the contrast between foreground and background is low and much of the lighter parts of the fish (such as its tail fin) are marked as background.

@@

\subsubsection{Fallback} The pixel analysis process occasionally fails,
e.g.~ flood-filling does not produce a large enough blob or
the bounding box adjustment does not terminate.
This can occur if certain flood fill operations behave unexpectedly, or if the image is too washed out or otherwise atypical for the thresholding process to work correctly. In the event this happens, the original mask and bounding box generated by \verb|detectron| is used for metadata generation.

\subsection{Metadata Generation}

The following metadata properties are generated from the methods
described above:
\verb|has_fish|, \verb|fish_count|, \verb|has_ruler|, \verb|ruler_bbox|, \verb|background.{mean,std}|, \verb|foreground.{mean,std}|, \verb|bbox|, \verb|mask|, \verb|score|, and \verb|has_eye|.
The remaining properties are computed as described below, with all the
metadata properties listed in Table \ref{tab:properties}.

\subsubsection{centroid and eye\_center}
Centroids are provided for the masks and bounding boxes generated by \verb|detectron|, and since we do not recalculate the mask of fish eyes we can use that value directly for \verb|eye_center|.

Since we recalculate the mask of the fish, we must recalculate its centroid as well. This can be done via
\begin{equation}
    (\bar{x}, \bar{y}) = (\mathrm{round}(\frac{M_{10}}{M_{00}}), \mathrm{round}(\frac{M_{01}}{M_{00}})),
\end{equation}
where \(M_{00}\) is the pixel area of the fish's blob, \(M_{10}\) is the sum of all the \(x\) values of blob pixels, and \(M_{01}\) is the sum of all the \(y\) values of blob pixels.
\subsubsection{side}
Determining which \verb|side| of the fish is visible is predicated on finding its eye. If an eye is found, the sign of the \(x\) component of the vector from the centroid of the fish to the centroid of the eye specifies which side is up: negative for left and positive for right. This assumes the fish was photographed vertically, which is essentially always the case for all image collections our group has worked on.
\subsubsection{primary\_axis and clock\_value}
The \verb|primary_axis| of a fish can be calculated by taking the covariance\todo{I could explain this in more detail if necessary. (DEB No, that's OK.} of its blob in \(x\) and \(y\), which yields its principle eigenvector.
The eigenvector can be directly assigned to the property. If an eye is
present, we ensure that \verb|primary_axis| points in the direction of the
eye in relation to the fish's centroid.

Our team encoded this information as a ``clock value'' between 1 and 12 when manually recording it. To convert \verb|principal_axis| to \verb|clock_value|, the sign of \(x\) and \(y\) on the principal axis are used to determine which Cartesian quadrant the fish angles into relative to its centroid. 
Depending on this quadrant, we dot product the principal axis with either \([-1,0]\), \([0,-1]\), \([1,0]\) or \([0,1]\), which correspond to 9, 6, 3 and ``0'' o'clock respectively. The resulting radian value is then converted to a polar displacement in clock value space, and added to the comparative clock value used in the dot product. This gives the fish's clock value from
0 to \(11.\overline{9}\). Before recording \verb|clock_value| in the output,
the value is rounded to the nearest integer, with a 0 final result replaced with 12.
\subsubsection{scale and length}
\(\frac{\mathrm{pixels}}{\mathrm{inch}}\) can be calculated by measuring the distance in pixels between the digits 2 and 3 (a 1 inch separation) found on the ruler by \verb|detectron|. Converting this to \(\frac{\mathrm{pixels}}{\mathrm{cm}}\) gives the \verb|scale| metadata property as reported in the output.

For the fish \verb|length| property, it is necessary to determine the furthest points from the centroid of the fish in each direction along its major
axis. Since fish are normally measured in a straight line from their snout down the middle of their trunk, every pixel of the fish blob is projected
onto the fish's major axis (as a line through its centroid).
The projection is done by finding the closest point on the
centroid--principal axis line from the pixel's location.
% intersecting the line through the point along the tangent of the principal axis and the aforementioned centroid--principal axis line.
After processing every pixel in the fish blob, computing the distance
between the two furthest projected points gives the length of the fish in pixels. Multiplying this distance by \verb|scale| gives the fish \verb|length| in centimeters.
\begin{comment}
\subsection{Consideration}
Just gonna throw some caveats in here, might be its own section might not
\begin{itemize}
    \item Program falls back to using detectron mask and bbox for metadata generation in the event the mask generation/pixel analysis fails. Failure is defined as never finding a cohesive flood filled mask that covers more than 10\% of the detectron bbox initially. Can provide a statistic on how often this occurs.
\end{itemize}
\end{comment}
\section{Results}
Technicians employed by our team have manually generated the 22 metadata properties deemed crucial to the overall BGNN project~\cite{Leipzig2021.01.28.428644} for a large number of INHS images. \(20,699\) total entries were created by 13 technicians that spanned \(8,398\) unique images. Of these \(8,398\) images, \(7,244\) were both not part of the \verb|detectron| training set and met our current admissibility criteria for \verb|detectron| and pixel analysis processing. We ran the metadata extraction program on these \(7,244\) images.
For the properties of image scale, fish length, and fish bounding boxes
(properties not manually generated),
a random sample of 50 specimens from the set of \(7,244\) were analyzed by
hand for comparison.

@@

\subsection{Fish Detection}
All images in the INHS dataset contain exactly one fish. For \(7,209\) of the specimen images, one fish was detected, a 99.5\% correct rate.
For 25 of the images, 2 fish were detected, 3 fish were detected for 3 images, and for 7 images no fish were found. In the case of greater than 1 fish, 9 of the 28 contained tags which overlapped the fish and were themselves labeled as a second fish. Of the remaining 17, \verb|detectron| erroneously labeled the fish as two separate fish objects, or labeled a subsection of the fish a second time.

\subsection{Ruler Detection}
For all but 2 of the \(7,244\) images \verb|detectron| was able to find the ruler, a nearly perfect correct rate.
\todo{Not sure if I can phrase it that way or not since I haven't looked at all the images to see if the ruler itself was found or something else was labeled a ruler} In 56 of the images, the ruler itself was found but the ``2'' and/or the ``3'' on it was not, so a scale calculation could not be performed, producing a 99.2\% success rate for being able to compute 
\verb|scale|.

@@

\subsection{Foreground/Background Contrast Results?}
\textbf{Todo}: It would be great to compute a histogram of foreground.mean -
bakground.mean for the all our samples.
Perhaps it would then be obvious how to break the images into low, medium
and high F/B contrast.  We could show a representative image from each
group.

@@

\begin{figure}[H]
  \centering
  \includegraphics[width=0.49\linewidth]{images/011196_pred}
  \includegraphics[width=0.49\linewidth]{images/011196_mask}
  \includegraphics[width=0.49\linewidth]{images/52399_pred}
  \includegraphics[width=0.49\linewidth]{images/52399_mask}
  \includegraphics[width=0.49\linewidth]{images/56885_pred}
  \includegraphics[width=0.49\linewidth]{images/56885_mask}
  \caption{Examples of masks and bounding boxes from detectron (left) and pixel analysis (right).}
\end{figure}
Fish bounding boxes were calculated for all \(7,237\) images in which a fish was found. All but 263 of these were generated via pixel analysis, with those 263 falling back to the original \verb|detectron| bounding box. Of the 50 images reviewed manually for these properties, all mask and bounding boxes were correctly placed on/around the location of the fish. However, a number of them lacked portions of lightly colored tails and/or fins. Specifically, 22 mask and bounding boxes covered the entire fish, 16 missed some of the tail, and 12 missed most or all of the tail.
